Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-24 paper/paper.qmd
update paper

---
title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.285 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

The structure of this paper is organized as follows: @sec-data provides
details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions and
specifications of our linear regression framework. In @sec-result, we
present our findings, emphasizing the key predictors of Harris's
support. Finally, @sec-discussion explores the implications of our
results and suggests potential directions for future research. @sec-a
provides external data detail, @sec-b provides model detail, @sec-c
provides an datasheet, and @sec-d provides an detailed introduction of
surveys, sampling and observational data in our paper.

@@

@fig-choromap is a choropleth map visualizes the average infant death
rate per 1,000 live births across U.S. states, with data categorized by
color intensity. States with higher infant death rates are shown in
lighter shades (yellow), while states with lower rates are represented
in darker shades (purple). States with missing data are indicated in
grey. The map highlights disparities in infant mortality rates, with
some states, particularly in the southeastern U.S., exhibiting notably
higher rates compared to others. States like Alabama and Mississippi
show the highest rates, while states such as California and New York
exhibit lower rates.

@@

# Model

@@

The model focuses on the infant death rate per 1,000 live births as the dependent variable and includes the following independent variables:\
	-	Treatment Group Indicator `abortion_illegal`: A binary variable indicating the group: 1 if abortion is illegal in the state. 0 if abortion remains legal.\
	-	Time Indicator `after_injunction`: A binary variable indicating the time period: 0 for the period before the injunction implemented. 1 for the period after the injunction implemented.
	-	Interaction Term`after_injunction` × `abortion_illegal`: The interaction term captures the additional change in the outcome (e.g., percentage of support, behavior, or health-related metrics) for regions where abortion is illegal after the injunction implemented, beyond any changes observed in regions where abortion remains legal.

To account for potential regional variations in infant death rates, we include state as a random effect in the model. This allows us to control for unobserved heterogeneity across states that might introduce bias. Additionally, we include age_of_mother, mothers_single_race, and year_of_death as control variables to adjust for demographic and temporal factors that could influence infant mortality rates, ensuring that the estimated effects are not confounded by these characteristics.

By isolating the impact of the abortion injunction through this interaction term, the model identifies changes in infant death rates that can be attributed to the policy intervention. Additional details and diagnostics supporting the model’s implementation are provided in [Appendix-@sec-model-details].

@@

DID is a causal inference method that estimates the impact of an intervention by comparing changes in outcomes over time between a treatment group and a control group. By assuming parallel trends in the absence of the intervention, the DID framework isolates the intervention’s effect through the interaction between time periods (pre- and post-intervention) and group membership. This makes it a robust method for evaluating policy changes and other significant events.

@@

Define $y_i$ as as the infant death rate per 1,000 live births.
$$y_i|\mu_i,\sigma \sim \mbox{Normal}(\mu_i, \sigma)\\
\text{First Model}: \mu_i = \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}+ \beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
\text{Second Model}:\mu_i = \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
\beta_0 \sim \mbox{Normal}(0, 2.5) \\
\beta_1 \sim \mbox{Normal}(0, 2.5) \\
\beta_2 \sim \mbox{Normal}(0, 2.5) \\
\beta_3 \sim \mbox{Normal}(0, 2.5) \\
\beta_4 \sim \mbox{Normal}(0, 2.5) \\
\beta_5 \sim \mbox{Normal}(0, 2.5) \\
\beta_6 \sim \mbox{Normal}(0, 2.5) \\
\gamma_j \sim \text{Normal}(0, \sigma_j^2)\\
\sigma \sim \mbox{Exponential}(1)$$

@@

The Bayesian framework enhances the DID approach by incorporating prior information and quantifying uncertainty in a probabilistic manner. Using Markov Chain Monte Carlo (MCMC) sampling, the model estimates the posterior distributions of the parameters, offering flexibility in handling complex relationships and uncertainties in the data.

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: false

@@

modelsummary::msummary(
  list(first_model,second_model),
  fmt = 3,coef_omit="state",notes = "Note. All models include a control for state and pollsters. The reference level of the state is Alaska, the reference level of population is likely voter, and the reference level of pollsters is Angus Reid.")
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The
discussion is a chance to show off what you know and what you learnt
from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what
your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

# Additional data details

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
check. This shows...

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

```{r}
plot(first_model, pars = "(Intercept)", prob = 0.95)
```

```{r}
prior_summary(second_model)
```

@@

plot(first_model, "rhat")
```

```{r}
plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```

```{r}
pp_check(second_model) +
  theme_classic() +
  theme(legend.position = "bottom")
```

>>>>

---
title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.285 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study examines the impact of the Dobbs decision and subsequent abortion bans on infant mortality rates in the United States, focusing on data from 2021 to 2022. Using a Difference-in-Differences (DID) approach, we analyze trends in states where abortion remained legal compared to those where it became illegal after June 2022. Our results indicate that states enforcing abortion bans experienced an increase of 0.285 infant deaths per 1,000 live births, with compounded effects observed during the post-injunction period. Maternal age and race also emerged as significant predictors, with Black mothers facing disproportionately higher infant mortality rates. These findings underscore the compounded public health implications of restrictive abortion policies and the urgent need for targeted interventions to address racial and demographic disparities. By utilizing recent, high-quality data and rigorous modeling, this study provides critical insights into the intersection of reproductive rights and infant health outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

The structure of this paper is organized as follows: @sec-data provides
details on the data sources and variables used in our analysis.
@sec-model explains the modeling approach, including the assumptions and
specifications of our linear regression framework. In @sec-result, we
present our findings, emphasizing the key predictors of Harris's
support. Finally, @sec-discussion explores the implications of our
results and suggests potential directions for future research.
[Appendix- @sec-data-details] provides external data detail, [Appendix-
@sec-model-details] provides model detail, @sec-c provides an datasheet,
and @sec-d provides an detailed introduction of surveys, sampling and
observational data in our paper.

@@

@@

## Random Variable

@fig-choromap is a choropleth map visualizes the average infant death
rate per 1,000 live births across U.S. states, with data categorized by
color intensity. States with higher infant death rates are shown in
lighter shades (yellow), while states with lower rates are represented
in darker shades (purple). States with missing data are indicated in
grey. The map highlights disparities in infant mortality rates, with
some states, particularly in the southeastern U.S., exhibiting notably
higher rates compared to others. States like Alabama and Mississippi
show the highest rates, while states such as California and New York
exhibit lower rates. \# Model

@@

@@

Difference-in-Differences (DID) is a causal inference method that
estimates the impact of an intervention by comparing changes in outcomes
over time between a treatment group and a control group. By assuming
parallel trends in the absence of the intervention, the DID framework
isolates the intervention’s effect through **the interaction between
time periods (pre- and post-intervention) and group membership**. This
makes it a robust method for evaluating policy changes and other
significant events.

### Assumption

The validity of the Difference-in-Differences (DID) approach in our
study relies on several key assumptions, which, if violated, could
affect the robustness and reliability of our results [@cunningham2021]:

#### Parallel Trends Assumption

The model relies on the assumption that, without the abortion
injunction, the treatment group (states where abortion became illegal)
and the control group (states where abortion remained legal) would have
followed similar trends in infant mortality rates. However, this
assumption can be challenged by differences in healthcare systems,
socioeconomic conditions, or pre-existing policy environments, which may
cause deviations from parallel trends.

#### Compositional Differences

In addition to parallel trends, compositional changes in repeated
cross-sectional data over time can influence results. For instance,
shifts in demographic factors such as maternal age or racial makeup
between pre- and post-intervention periods may confound the observed
changes in infant mortality rates. Including control variables helps
adjust for these confounders, ensuring that the estimated effect
reflects the policy intervention rather than changes in population
composition.

#### Long-term Effects vs. Reliability

The length of the analysis period further complicates the reliability of
results. While longer periods allow for capturing long-term effects,
they also increase the likelihood of unrelated factors, such as new
policies or healthcare advancements, influencing the outcome. Over time,
some states in the control group may also implement similar abortion
policies, diluting the treatment effect. Conversely, shorter analysis
periods

#### Functional Form Dependence

Finally, the model assumes that the functional relationship between
predictors and outcomes is correctly specified. If this assumption is
violated, differences in outcomes between treatment and control groups
may stem from misspecified relationships rather than the intervention
itself. This highlights the importance of carefully assessing the
model’s structure and assumptions to ensure valid results.

## Model Parameters

The model focuses on the infant death rate per 1,000 live births as the
dependent variable and includes the following independent variables:

-   **Treatment Group Indicator `abortion_illegal`**: A binary variable
    indicating the group: 1 if abortion is illegal in the state. 0 if
    abortion remains legal.

-   **Time Indicator `after_injunction`**: A binary variable indicating
    the time period: 0 for the period before the injunction implemented.
    1 for the period after the injunction implemented.

-   **Interaction Term`after_injunction` × `abortion_illegal`**: The
    interaction term captures the additional change in the outcome
    (e.g., percentage of support, behavior, or health-related metrics)
    for regions where abortion is illegal after the injunction
    implemented, beyond any changes observed in regions where abortion
    remains legal.

Additionally, we include these control variables to adjust for
demographic and temporal factors that could influence infant mortality
rates, ensuring that the estimated effects are not confounded by these
characteristics:

-   **`age_of_mother`**: The age of the mother at the time of birth,
    represented as a continuous variable.

-   **`mothers_single_race`**: Self-identified race of the mother,
    represented as a categorical variable.

-   **`year_of_death`**: The year in which the infant death occurred,
    represented as a continuous variable.

Finally, we include a random variable:

-   **`state`**: Mother's residence state when the infant born as a
    random effect in the model.

By isolating the impact of the abortion injunction through this
interaction term `after_injunction`×`abortion_illegal`, our model
identifies changes in infant death rates that can be attributed to the
policy intervention. Additional details and diagnostics supporting the
model’s implementation are provided in [Appendix- @sec-model-details].

@@

Define $y_i$ as as the infant death rate per 1,000 live births, and
$\gamma_j$ as the random state effect. $$
\begin{aligned}
y_i|\mu_i,\sigma &\sim \mbox{Normal}(\mu_i, \sigma)\\
\text{First Model}: \mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}\\&+\beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
\text{Second Model}:\mu_i &= \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i\\&+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\&+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
\beta_0 &\sim \mbox{Normal}(0, 2.5)\\
\beta_1 &\sim \mbox{Normal}(0, 2.5)\\
\beta_2 &\sim \mbox{Normal}(0, 2.5)\\
\beta_3 &\sim \mbox{Normal}(0, 2.5)\\
\beta_4 &\sim \mbox{Normal}(0, 2.5)\\
\beta_5 &\sim \mbox{Normal}(0, 2.5)\\
\beta_6 &\sim \mbox{Normal}(0, 2.5)\\
\gamma_j&\sim \text{Normal}(0, \sigma_j^2)\\
\sigma &\sim \mbox{Exponential}(1)
\end{aligned}$$

@@

For our analysis, we use separate Bayesian regression models to estimate
the infant death rates in states where abortion injunctions were
implemented and in states where abortion remains legal. We use the
interaction term between after_injunction and abortion_illegal to
measure the causal effect of the abortion injunction. Both models share
the same predictors to ensure consistency and comparability in capturing
the determinants of infant mortality.

For one of our models, we include control variables (`age_of_mother`,
`mothers_single_race`, and `year_of_death`) to adjust for confounders
that could influence infant death rates, addressing the assumption of
compositional differences in the DID approach and ensuring unbiased
estimates of the intervention’s effect. In the other model, we exclude
these controls to focus on the overall intervention impact, avoid
overfitting, and prioritize simplicity when variability is adequately
accounted for by the design or random effects. Thus, our decision to fit
two models is to compare the outcomes: one adjusts for confounders to
refine the estimates, while the other focuses on the direct effect of
the intervention.

Our model includes `state` as a random effect to account for regional
variations in infant mortality that could introduce bias. We use a
random effect rather than a fixed effect to capture variability across
states without estimating specific effects for each one. This approach
models unobserved heterogeneity, such as differences in healthcare
systems and socioeconomic conditions, using a single variance term,
which avoids overfitting and ensures efficiency and generalizability. By
adjusting for unmeasured state-specific factors, the model provides
robust and reliable findings while balancing parsimony and
interpretability, making it well-suited for our analysis.

We choose the Bayesian framework since it enhances the DID approach by
incorporating prior information and quantifying uncertainty in a
probabilistic manner. In our Bayesian models, we use weakly informative
priors from the `rstanarm` package, including a Normal prior ($\mu=0$
and $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1)
for the auxiliary parameter, with `autoscale=TRUE` to dynamically adjust
priors based on data scale. This enhances convergence and ensures priors
are calibrated to predictor ranges without dominating the data. The
model assumes plausible parameter centering, random effects capture
unobserved heterogeneity, and the DID framework relies on parallel
trends between groups. However, limitations include potential bias from
misspecified priors, computational challenges for hierarchical models,
and sensitivity to data sparsity or imbalances. This model may not be
suitable if group-level effects are negligible, parallel trends are
violated, or data is insufficient, but when applied appropriately, it
provides robust and interpretable causal estimates.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

@@

modelsummary::msummary(
  list(first_model,second_model),
  fmt = 3,coef_omit="state",notes = "Note. All models include a random variable for state. The reference level of age_of_mother is 15-19.The reference level of monthers_single_race is Asian.")
```

The results from the two Bayesian models are summarized in
@tbl-modelresults. The interaction term between after_injunction and
abortion_illegal is positive and statistically significant (0.111 in
Model 1 and 0.280 in Model 2), indicating a change in infant mortality
rates associated with the combination of abortion bans and the
post-injunction period.

For the independent effects, states where abortion is illegal show
higher infant mortality rates, with coefficients of 3.365 in Model 1 and
1.400 in Model 2. The post-injunction period alone is associated with
coefficients of -0.525 in Model 1 and -0.940 in Model 2, indicating a
decrease in infant mortality during this time.

Demographic variables also exhibit notable associations. Maternal age is
negatively associated with infant mortality, with coefficients for age
groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, respectively. The
coefficient for Black mothers in Model 2 is 6.66, reflecting higher
infant mortality compared to other racial groups.

The models performed well, with low RMSE values and high $R^2$ values,
indicating that the predictors account for a substantial portion of the
variance in infant mortality rates.

@@

## Key Findings and Implications

The analysis reveals distinct patterns in infant mortality rates across
various policy and demographic factors. In states where abortion
remained legal, the post-injunction period (after_injunction1) is linked
to reductions in infant mortality rates by 0.525 and 0.940 deaths per
1,000 live births in Models 1 and 2, respectively. However, states with
abortion bans (abortion_illegal1) exhibit significantly higher infant
mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
deaths in Model 2. The DID interaction term (after_injunction1 ×
abortion_illegal1) reveals that abortion bans during the post-injunction
period result in an additional increase of 0.111 deaths in Model 1 and
0.280 deaths in Model 2. Demographic factors also play a critical role,
with older maternal age (25–34 years) reducing infant mortality rates,
while younger mothers (15–19 years) face elevated risks. Black mothers
experience disproportionately higher infant mortality rates, with an
increase of 6.660 deaths per 1,000 live births compared to Asian
mothers. These findings collectively emphasize the compounded effects of
policy and demographic disparities on infant health outcomes.

## Public Health Implications of Restrictive Abortion Laws

The findings highlight the significant public health consequences of
abortion bans in the United States. The observed increase of 0.285
infant deaths per 1,000 live births in states enforcing abortion bans
after the Dobbs decision underscores the detrimental impact of
restrictive abortion policies. These increases often reflect systemic
issues such as inadequate prenatal care, gaps in healthcare access, and
underlying socioeconomic inequities. States with pre-existing healthcare
challenges may experience amplified effects, suggesting an urgent need
for comprehensive policy interventions to address these disparities.
Tackling the root causes of these health disparities—such as improving
prenatal care, expanding healthcare access, and addressing socioeconomic
inequalities—will be essential for mitigating the unintended
consequences of restrictive abortion legislation.

## Racial and Socioeconomic Disparities

The results reveal stark racial disparities in infant mortality rates,
with Black mothers experiencing significantly higher rates compared to
other racial groups. These disparities are exacerbated by restrictive
abortion policies, which disproportionately affect already vulnerable
populations. The intersection of systemic healthcare inequities,
socioeconomic challenges, and restrictive legislation creates compounded
risks for marginalized communities, leading to worse health outcomes.
Addressing these inequities requires targeted policies that ensure
equitable access to maternal and infant healthcare, particularly in
communities most impacted by these disparities. Such policies should
prioritize investments in healthcare infrastructure, culturally
competent care, and programs that support at-risk populations.

## Integration of Policy and Healthcare Systems

The significant interaction between abortion legality and the
post-injunction period illustrates the compounded impact of restrictive
abortion policies on infant mortality rates. These findings underscore
the need for integrated healthcare systems that can address the
consequences of high-risk pregnancies, improve prenatal care, and ensure
equitable access to medical services. Policymakers must consider the
broader public health implications of abortion bans and implement
strategies that safeguard maternal and infant health. This includes
enhancing healthcare delivery in underserved areas, investing in
community-based interventions, and ensuring that restrictive policies
are accompanied by measures to mitigate their public health impact.

## Data and Temporal Limitations

A key limitation of our analysis is the temporal and demographic scope
of the dataset. The data includes only infant deaths under 1 year of age
occurring within the United States to U.S. residents during 2021–2022.
This timeframe does not capture the full year of infant mortality
outcomes following the Dobbs decision, as infants who died in 2023 are
not included. Furthermore, the dataset’s focus on infants under 1 year
of age may exclude cases that reflect longer-term health impacts. These
limitations could lead to an underestimation of the full effects of
abortion bans. Future research should incorporate more comprehensive and
updated datasets to provide a clearer and more complete analysis of the
long-term trends and implications of restrictive abortion policies.

## Weaknesses and Future Directions

While this study provides valuable findings, several limitations should
be addressed in future research. The analysis relies on aggregate
state-level data, which may obscure localized effects and intra-state
disparities. Unmeasured confounders, such as economic conditions,
healthcare infrastructure differences, or state-specific policies, could
also influence the results. Moreover, the short timeframe of the dataset
limits the ability to assess the long-term effects of abortion policies.
Future research should integrate more granular data, such as
county-level analyses, and extend the study period to capture
longer-term impacts. Qualitative studies focusing on the lived
experiences of affected populations can complement quantitative
analyses, offering a deeper understanding of the broader implications of
restrictive abortion laws. These efforts will support the development of
evidence-based policies aimed at reducing disparities and improving
maternal and infant health outcomes.

@@

# Additional data details {#sec-data-details}

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
check for the first model. This shows...

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["First Model Posterior prediction check", "First Model Comparing the posterior with the prior","Second Model Posterior prediction check", "Second Model Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")
pp_check(second_model) +
  theme_classic() +
  theme(legend.position = "bottom")
posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
posterior_vs_prior(second_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
prior_summary(first_model)
prior_summary(second_model)
```

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-area
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["First Model Posterior prediction check", "First Model Comparing the posterior with the prior","Second Model Posterior prediction check", "Second Model Comparing the posterior with the prior"]

plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```

@@

plot(first_model, "rhat")

plot(second_model, "trace")

plot(second_model, "rhat")
```

# Idealized Methodology for A Survey-Based Qualitative Studies {#sec-survey}

In our study, we examine the impact of abortion bans on infant mortality
rates in the United States following the Dobbs decision, employing both
observational data and survey-based research. Observational data provide
quantitative insights, revealing trends and associations across large
populations. However, they often fall short in addressing underlying
mechanisms due to limitations such as confounding factors and challenges
in establishing causality.

To address these gaps, in this appendix we propose incorporating surveys to gather
qualitative data, which provide rich, detailed accounts of individual
experiences and contextual factors. Qualitative methods allow for a
deeper understanding of how restrictive abortion policies affect
maternal health, access to healthcare, and infant outcomes, capturing
nuances that quantitative analysis alone cannot.

By combining quantitative and qualitative approaches, this study
utilizes the strengths of both methods. Quantitative data offer
generalizable findings and statistical rigor, while qualitative data
provide depth and context to interpret those findings. Together, they
create a more comprehensive framework for understanding the complex
relationship between restrictive abortion policies, maternal health, and
infant outcomes, enabling the development of nuanced, evidence-based
interventions and policy recommendations.

## Introduction

Restrictive abortion laws may exacerbate maternal stress, anxiety, or
depression—especially in cases of unintended or high-risk
pregnancies—potentially undermining maternal health and prenatal care.
Research indicates that maternal mental health is a critical determinant
of infant health outcomes, as conditions such as chronic stress,
depression, or anxiety during pregnancy can negatively impact prenatal
behaviors, access to healthcare, and overall maternal well-being. These
psychological challenges, compounded by societal pressures and systemic
barriers, can significantly increase the risk of adverse infant
outcomes, including infant mortality. The survey will focus on mothers
unable to access abortion services who subsequently experienced the
death of their infant, documenting their experiences to better
understand these interconnected factors.

To provide a comprehensive perspective, the study will also involve
healthcare providers and support organizations, capturing insights into
systemic challenges and the broader impacts of abortion bans on family
healthcare. By combining these perspectives with mothers’ lived
experiences, the survey aims to inform evidence-based interventions
addressing maternal psychological and healthcare needs and mitigating
the negative consequences of restrictive abortion policies.

This dual approach, integrating quantitative and qualitative methods,
seeks to uncover the complex relationship between restrictive abortion
laws, maternal mental health, and infant outcomes, forming the basis for
effective policy recommendations.

## Objective

The objective of this study is to investigate the psychological, social,
and systemic impacts of restrictive abortion laws on maternal health and
infant outcomes. By focusing on mothers unable to access abortion
services who experienced the death of their infant, the study aims to
document lived experiences, explore the role of maternal stress and
healthcare barriers in adverse outcomes, and gather perspectives from
healthcare providers and support organizations. Understanding how
maternal mental health directly affects prenatal care decisions, access
to healthcare, and overall maternal behavior is essential to addressing
the systemic factors contributing to higher infant mortality rates. The
findings will guide the development of evidence-based strategies to
improve maternal mental health, enhance healthcare access, and shape
policy responses to restrictive abortion laws.

## Sampling Approach

In this analysis, we will use non-probability Respondent-driven sampling to finding respondents. Respondent-driven sampling (RDS), developed by @heckathorn1997, is ideal for studying hidden populations lacking a sampling frame and facing potential stigma or harm. Unlike snowball sampling, RDS offers dual compensation for participation and recruitment, with recruitment occurring peer-to-peer via social networks. This method maintains confidentiality, reduces bias through statistical adjustments, and is particularly effective for sensitive topics, such as abortion, by leveraging trust within communities.

We use Respondent-Driven Sampling (RDS) for our survey because it is well-suited for reaching hidden or stigmatized populations, such as mothers who were unable to access abortion services and experienced the death of their infant. These individuals may be difficult to identify through conventional methods due to the sensitive nature of the topic and the lack of a clear sampling frame.

RDS utilizes peer networks, allowing participants to recruit others they trust, fostering higher participation rates in populations that may distrust external researchers. This approach helps maintain confidentiality and reduces the risk of social stigma. Additionally, RDS provides a structured framework for statistical adjustments, mitigating biases inherent in network-based recruitment, and improving the representativeness of the sample. These advantages make RDS an effective and ethical choice for gathering insights into the psychological, social, and systemic impacts of restrictive abortion laws.

## Target Population

Our target population is mothers who sought but were unable to access abortion services due to restrictive laws and experienced the death of their infant within the first year of life in states with abortion bans following the Dobbs decision.

## Sample frame

Given the sensitive nature of our study, identifying and reaching the initial participants poses a significant challenge. To address this, we aim to collaborate with support groups and advocacy organizations like Abortion Rights Advocacy Groups. These groups play a critical role in connecting researchers with individuals who are otherwise difficult to reach, particularly for sensitive topics like the impact of restrictive abortion laws on maternal and infant health.

Therefore, the sample frame of our study will be mothers who sought but were unable to access abortion services due to restrictive laws and experienced the death of their infant within the first year of life in states with abortion bans following the Dobbs decision and are reachable and can be contacted by Abortion Rights Advocacy Groups.

## Sample

We plan to survey 10 respondents for this study. Our sample will consist of mothers who fit the defined sample frame—those who sought but were unable to access abortion services due to restrictive laws, experienced the death of their infant within the first year of life, and are reachable through Abortion Rights Advocacy Groups. These mothers must also be willing to participate in and answer our survey questions, ensuring the reliability and depth of the collected data.

## Ethical Concerns

This study addresses sensitive and emotionally charged topics, necessitating a careful and ethical approach to protect participants’ well-being. We acknowledge the potential psychological distress that discussing experiences of unintended pregnancies, restrictive abortion laws, and infant loss may cause. To mitigate this, the survey will include trigger warnings, and participants will have the option to skip questions or withdraw at any time without penalty. All data collection will adhere to strict confidentiality protocols, ensuring that participants’ identities remain anonymous and their responses are securely stored. Recruitment through trusted advocacy groups will ensure that participants feel safe and supported throughout the process. Additionally, we will provide resources for mental health support to any participant who may experience discomfort as a result of participating in the study. This approach underscores our commitment to conducting ethical, respectful, and responsible research.

## Proposed Survey Design
